### 2017-08-31 paper.tex
:construction: Initiation phase

>>>>

\section{Mechanisms of granular column collapse in fluid}
The difference in the runout behaviour between the dense and loose granular columns suggests difference in runout mechanism. The collapse of a granular column involves three stages: an initiation stage characterised by a distinct failure surface, a runout phase characterised by horizontal acceleration which involves formation of eddies, and the final settlement phase. The mechanisms observed in each phase are discussed next.

\subsection{Initiation phase}
Comparing the initial evolution of runout between the loose and the dense cases ($t = 0$ to $3\tau_c$), see~\cref{fig:slope_loose_5,fig:dense-s5-r09}, the loose granular columns evolve faster than their dense counterpart. When a granular material is sheared in the submerged conditions, it generates negative pore water pressure initially due to the undrained conditions. In the undrained conditions, the dilation movement (volume expansion) of the granular assembly by shearing self-generates negative excess pore pressure inside the pores. The fluid inside the pores does not have time to seep in or out from the free water outside the column by the internally generated pressure gradient between the pore space and the outside free water. The negative excess pore pressures produce larger interparticle forces by bringing the grains together. Macroscopically this results in increase in the mean effective stress and provides large shear resistance temporarily until the negative excess pore pressure dissipates with time. Large negative pore pressures were observed for the dense cases. The length of the failure plane increases with the slope angle, thereby increasing the volume of body with negative pore pressure that needs to be dissipated before the column starts to move. Overcoming the region of large negative pore pressure translates to smaller kinetic energy and eventually a shorter runout distance for the dense granular columns.

The degree of dilation upon shearing is smaller in the loose granular column case than the dense case. Positive pore pressures can be observed along the failure plane. Positive pore pressures imply that the grains are pushing apart by increasing pore pressure during the initial phase of the collapse. The intergranular forces (or effective stress) are reduced and consequently the shear resistance decreases. This results in faster initiation and runout behaviour for the loose granular columns compared to the dense granular columns~\citep{Kumar2017}.
